---
title: "HW5"
author: "Kevin Mack"
date: "2/17/2016"
output: pdf_document
---

Question 1:
-
I examined the dataset and selected a relatively common species, T. tuberculata and a relatively rare species, P. elegans. I examined the relationship between the abundance of each species and space based on the BCI dataset. 

```{r, echo=FALSE}
library(vegan)
library(nlme)
data(BCI)
BCI_xy = data.frame(x = rep(seq(625754, 626654, by=100), each=5), 
                    y = rep(seq(1011569,  1011969, by=100), len=50))
```

```{r}
boxplot(BCI)
plot(BCI$Trichilia.tuberculata)# abundant
plot(BCI$Platypodium.elegans) # rare, but enough for me to work with
BCI_sub=subset(BCI, select= c(Trichilia.tuberculata, Platypodium.elegans))
BCI_lots=BCI$Trichilia.tuberculata
BCI_few=BCI$Platypodium.elegans

boxplot(BCI_sub)
```

Common Species: T. tuberculata is abundant and widely distributed. This species does not exhibit any form of spatial dependence. Biologically, this makes sense, a species that occurs commonly and in high abundance in a given habitat is not likely to be dependent on any specific environmental variables within that habitat

```{r}
#visual summary of abundance
col_brks = hist(BCI_lots, plot=F)$breaks
col_indices = as.numeric(cut(BCI_lots, col_brks))
cols = rev(terrain.colors(length(col_brks)))
plot(BCI_xy, cex=2, pch=19, col=cols[col_indices], main="Spatial Distribution of T. tuberculata ")

#Determine distances
lots_dist = dist(BCI_lots)
xy_dist = dist(BCI_xy)
max_dist = max(xy_dist) / 2

# plot result
plot(xy_dist, lots_dist)
abline(lm(lots_dist ~ xy_dist), lwd=3, col='red') #positive line is pretty level
lines(lowess(xy_dist, lots_dist), lwd=3, col='pink')
abline(v = max_dist, col='red', lwd=3, lty=2) #line is fairly level, certianly levels off with distance. Abundance of this species does not continue to increase with distance

#Compute Correlation
lots_cor = cor(xy_dist, lots_dist)
lots_cor # weakly correlated

#mantel test
lots_mantel = mantel(xy_dist, lots_dist)
lots_mantel # observed spatial patterns not significantly different from random.
```

Rare Species: P. elegans is patchy in distribution and more abundant in some patches than others. This species does appear to be spatially dependent, as it's distribution pattern is significantly different from random (p<0.01). P. elegans is a large tree whose spatial distribution is likely dependent on it's ability to out compete other trees for sunlight, potentially explaining its patchy, spatially dependent distribution.


```{r}
#visual summary of abundance
col_brks = hist(BCI_few, plot=F)$breaks
col_indices = as.numeric(cut(BCI_few, col_brks))
cols = rev(terrain.colors(length(col_brks)))
plot(BCI_xy, cex=2, pch=19, col=cols[col_indices], main="Spatial Distribution of P. elegans")

#Determine distances
few_dist = dist(BCI_few)

# plot result
plot(xy_dist, few_dist)
abline(lm(few_dist ~ xy_dist), lwd=3, col='red') #positive line is pretty level
lines(lowess(xy_dist, few_dist), lwd=3, col='pink')
abline(v = max_dist, col='red', lwd=3, lty=2) #line has distinct positive slope which does not appear to level off, abundance of this species increases with distance (?)

#Compute Correlation
few_cor = cor(xy_dist, few_dist)
few_cor # correlated

#mantel test
few_mantel = mantel(xy_dist, few_dist)
few_mantel # observed spatial patterns are significantly different from random.
```

Question2 :
-
I developed two GLMs to predict the abundance of a species of interest, D. standleyi  as a function of the abundance of other species.
```{r}
#Structuring data for the models
#thanks to Tom for this workaround

sp_ids = c("Cordia.lasiocalyx", "Hirtella.triandra","Picramnia.latifolia", "Quassia.amara",        "Tabernaemontana.arborea", "Trattinnickia.aspera","Xylopia.macrantha")
response = BCI$Drypetes.standleyi

modeling_dat = data.frame(response,BCI[,which(colnames(BCI) %in% sp_ids)],BCI_xy)

```

Model 1: Picking the species T. arborea, I constructed a general linear model to predict the abundance of our species of interest, D. standleyi. I also examined spatial dependence of the model residuals The model. Using only a single species and without accounting for spatial dependence, does not do a very adequate job of explaining abundance of the species of interest. Adding a spatial error term (in the form of Variograms) altered the model coefficients and generally improved the fit. Of the spatial error terms applied the rational quadratic model provided the best fit.


```{r}
sp_e = modeling_dat$Tabernaemontana.arborea #cause I dont want to type out Tabernaemontana.arborea a bajillion times, even though I just typed it out an extra time to say that. Actually, now its fun: Taverninmontana.arborday

mod1 = gls(response~sp_e, data=modeling_dat)
summary(mod1)

plot(Variogram(mod1, form= ~ x + y, nugget=T)) #dat nugget

#residuals
res = residuals(mod1)
plot(dist(modeling_dat[, c('x', 'y')]), dist(res))
lines(lowess(dist(modeling_dat[, c('x', 'y')]), dist(res)), col='red', lwd=2)
abline(v = max_dist, col='red', lwd=3, lty=2)

#fit of error model: expodential
mod1_exp = update(mod1, corr=corExp(form=~x + y))
summary(mod1_exp)
plot(Variogram(mod1_exp, maxDist = max_dist)) #yeah, this is bad
plot(Variogram(mod1_exp, resType = 'normalized', maxDist = max_dist))

#fit of error model: rational quadratic
mod1_rat = update(mod1_exp, corr=corRatio(form=~x+y, nugget=T))
summary(mod1_rat)
plot(Variogram(mod1_rat, maxDist = max_dist))
plot(Variogram(mod1_rat, resType = 'n',  maxDist = max_dist)) #much better

#fit of error model: Spherical
mod1_spher = update(mod1_rat, corr=corSpher(form=~x+y, nugget = T))
summary(mod1_spher)
plot(Variogram(mod1_spher, maxDist = max_dist))
plot(Variogram(mod1_spher, resType = 'n',  maxDist = max_dist)) #rational quadratic better

#fit of error model: Gaussian
mod1_gaus = update(mod1_spher, corr=corGaus(form=~x+y, nugget=T))
summary(mod1_gaus)
plot(Variogram(mod1_gaus, maxDist = max_dist))
plot(Variogram(mod1_gaus, resType = 'n',  maxDist = max_dist)) #also pretty good


anova(mod1, mod1_exp, mod1_rat, mod1_spher, mod1_gaus)
#rational quadratic model has lowest AIC, barely
```

Model 2: To develop a model to better predict the abundance of our species of interest, additional predictor species were included in a second GLM. This model better predicts the abundance of our species of interest. Again, the addition of an spatial error term was able to provide some improvement of the fit. In this instance, exponential spatial error term provided the best fit. 

```{r}
mod2=gls(response~Cordia.lasiocalyx+Hirtella.triandra+Picramnia.latifolia+Quassia.amara+Tabernaemontana.arborea+Trattinnickia.aspera+Xylopia.macrantha, data=modeling_dat)
summary(mod2)


plot(Variogram(mod2, form=~x+y))

#residuals
res2 = residuals(mod2)
plot(dist(modeling_dat[, c('x', 'y')]), dist(res2))
lines(lowess(dist(modeling_dat[, c('x', 'y')]), dist(res2)), col='red', lwd=2)
abline(v = max_dist, col='red', lwd=3, lty=2)

#fit of error model: expodential
mod2_exp = update(mod2, corr=corExp(form=~x + y))
summary(mod2_exp)
plot(Variogram(mod2_exp, maxDist = max_dist,  data=modeling_dat)) #yeah, this is bad
plot(Variogram(mod2_exp, resType = 'normalized', maxDist = max_dist,  data=modeling_dat))

#fit of error model: rational quadratic
mod2_rat = update(mod2_exp, corr=corRatio(form=~x+y, nugget=T))
summary(mod2_rat)
plot(Variogram(mod2_rat, maxDist = max_dist,  data=modeling_dat))
plot(Variogram(mod2_rat, resType = 'n',  maxDist = max_dist,  data=modeling_dat)) 

#fit of error model: Spherical
mod2_spher = update(mod2_rat, corr=corSpher(form=~x+y, nugget = T))
summary(mod2_spher)
plot(Variogram(mod2_spher, maxDist = max_dist,  data=modeling_datt))
plot(Variogram(mod2_spher, resType = 'n',  maxDist = max_dist,  data=modeling_dat)) 

#fit of error model: Gaussian
mod2_gaus = update(mod2_spher, corr=corGaus(form=~x+y, nugget=T))
summary(mod2_gaus)
plot(Variogram(mod2_gaus, maxDist = max_dist))
plot(Variogram(mod2_gaus, resType = 'n',  maxDist = max_dist)) 


anova(mod2, mod2_exp, mod2_rat, mod2_spher, mod2_gaus)
#exponential model has lowest AIC, barely. 


```

